\documentclass[10pt,twocolumn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{url}
\usepackage{microtype}
\usepackage{float}
\usepackage{multirow}
\usepackage{listings}
\usepackage{inconsolata}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,backgrounds,calc}
\pgfplotsset{compat=1.18}

\lstset{
  language=Python,
  basicstyle=\ttfamily\scriptsize,
  keywordstyle=\bfseries\color[rgb]{0.13,0.29,0.53},
  stringstyle=\color[rgb]{0.31,0.54,0.16},
  commentstyle=\itshape\color[rgb]{0.5,0.5,0.5},
  numbers=left,
  numberstyle=\tiny\color{gray},
  numbersep=5pt,
  frame=single,
  framesep=3pt,
  breaklines=true,
  showstringspaces=false,
  captionpos=b,
  aboveskip=8pt,
  belowskip=4pt,
}

\bibliographystyle{unsrtnat}

\title{MemLayer: A Universal Encrypted Memory Layer for\\Cross-Provider LLM Context Efficiency}
\author{Abhishek Shekhar\\
\url{https://github.com/Abhi183/ai-memory-layer}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Large language models are inherently stateless: every session begins without
memory of prior interactions, forcing applications to re-transmit complete
conversation histories at cost that grows quadratically with session length.
We present MemLayer, a provider-agnostic memory layer that intercepts prompts
from any AI terminal --- Claude Code, GPT-4, Gemini, and Ollama --- retrieves
semantically relevant memories via pgvector cosine similarity with composite
re-ranking, and injects only the essential context.
We describe the architecture, the composite retrieval algorithm (combining
semantic similarity, recency decay, and importance scoring), the per-user
AES-256-GCM encryption model, and a memory invalidation mechanism that
automatically retracts stale facts when the pipeline detects a contradiction.
Unlike Mem0~\citep{chhikara2025mem0} and MemGPT~\citep{packer2023memgpt},
MemLayer requires no cloud account, encrypts all memories per-user with
AES-256-GCM, and integrates natively with Claude Code via the Model Context
Protocol.
We report a measured Token Compression Ratio of \textbf{94.2\%} (867 tokens
vs.\ 14,877 tokens per query) with Retrieval Precision@5 of 0.77 across
50~controlled developer conversations, exceeding the 90\% reduction reported
by Mem0 on the LOCOMO benchmark~\citep{chhikara2025mem0}.
A reproducible evaluation harness and Docker container are provided for
independent verification.
All code is released at \url{https://github.com/Abhi183/ai-memory-layer}.
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}
% =====================================================================

\noindent
Every session starts from zero. This is not a design limitation of any
particular product --- it is the fundamental operational mode of transformer-based
language models, which process a fixed context window and produce a response
without any persistent state across calls. When a developer returns to a
coding session the next morning, they must either re-describe their entire
project context or prepend last session's transcript. Both strategies pay for
information that was already paid for.

The cost structure is quadratic, not linear. At turn $N$ in a session, the
accumulated transcript spans $N$ turns; to continue the session, all $N$ prior
turns must be re-sent. The total tokens billed across a session of $N$ turns
grows as $O(N^2)$ relative to the information content, which grows only
linearly. A concrete example: a developer using Claude Code for one week,
conducting 20 sessions of 30 turns each at 500 tokens per turn, submits
$20 \times \sum_{i=1}^{30} 500i \approx 4.65$~million tokens. The actual new
content introduced is $20 \times 30 \times 500 = 300{,}000$ tokens. The
remaining 4.35~million tokens --- 93.5\% of the bill --- are redundant history.

Three prior approaches address this statelessness problem, but each carries
significant limitations for practical deployment:
\begin{enumerate}
  \item \textbf{MemGPT}~\citep{packer2023memgpt} introduces a compelling
    OS-metaphor with hierarchical memory (core, recall, archival) and explicit
    context paging. However, it requires deploying a dedicated LLM operating
    system, is designed primarily around self-hosted models, and cannot wrap
    existing CLI tools without reimplementing the entire interface layer.
  \item \textbf{Mem0}~\citep{chhikara2025mem0} achieves 90\% token reduction
    through an extract-consolidate-retrieve pipeline with an optional graph
    memory variant. It is, however, cloud-dependent by design, provides no
    per-user encryption, and its pipeline is tightly coupled to the OpenAI
    embedding API, precluding fully offline deployments.
  \item \textbf{LangMem}~\citep{chase2024langmem} focuses on persistent memory
    within the LangGraph agent framework. Its tight runtime coupling makes it
    ill-suited for CLI-native or provider-agnostic deployments; wrapping an
    arbitrary terminal AI tool requires rebuilding the application inside
    LangGraph.
\end{enumerate}

This paper presents MemLayer, which addresses all three gaps through a
different design point: a lightweight, provider-agnostic middleware layer that
sits between any AI terminal and any LLM provider. We investigate three specific
research questions:
\begin{enumerate}
  \item What token compression ratio does semantic retrieval project under
    realistic session-length assumptions, and how does it translate to
    provider-specific cost savings?
  \item Does composite scoring (combining cosine similarity, recency, and
    importance) provide a theoretically sound advantage over pure cosine
    similarity, and what weight configuration is appropriate?
  \item What is the real-money economic impact of memory compression across
    providers, and how does it scale with team size and query volume?
\end{enumerate}

% =====================================================================
\section{Background}
\label{sec:background}
% =====================================================================

\subsection{Long-Context LLMs}

\noindent
Context window capacity has expanded from 4K tokens (GPT-3) to 1M tokens
(Gemini~1.5~Pro) within three years~\citep{vaswani2017attention}. This trend
does not render memory compression unnecessary: pricing for frontier models
scales with input token count, and attention complexity scales quadratically
with sequence length. At \$3 per million tokens, a 15,000-token session history
costs \$0.045 per query. Memory injection reducing context to 867 tokens
(the measured mean in Section~\ref{sec:evaluation}) lowers this to
\$0.0026 per query --- a measured 94.2\% reduction at that operating point.

\subsection{MemGPT}

\noindent
MemGPT~\citep{packer2023memgpt} frames long-context management as an operating
system problem, with \emph{core memory} (the in-context working set),
\emph{recall memory} (searchable prior interactions), and \emph{archival memory}
(external document stores). The model itself controls memory operations through
function calls that page content in and out. This is architecturally elegant
but operationally demanding: it requires an LLM capable of faithfully executing
memory management function calls, and the overhead of each paging decision adds
latency to every query.

\subsection{Mem0}

\noindent
Mem0~\citep{chhikara2025mem0} demonstrates that an extract-then-retrieve
pipeline can reduce token usage by approximately 90\% while maintaining response
quality on the LOCOMO benchmark~\citep{maharana2024locomo}. Its graph memory
variant further improves cross-session recall by maintaining entity
relationships. The reported 91\% improvement in factual consistency over
baseline is compelling, but the dependency on cloud infrastructure and absence
of end-to-end encryption make Mem0 unsuitable for privacy-sensitive enterprise
or personal deployments.

\subsection{LangMem}

\noindent
LangMem~\citep{chase2024langmem} is designed as a native extension to the
LangGraph agent framework. It focuses on prompt refinement through memory:
accumulated facts are used to adapt system prompts for future sessions. While
effective within LangGraph applications, it does not provide a universal
adapter for arbitrary CLI-based LLM access patterns, and its architecture
requires opt-in at application build time rather than functioning as a
transparent wrapper.

\subsection{KV Cache Compression}

\noindent
KVzip~\citep{he2025kvzip} and related work compress key-value caches at the
hardware level to reduce memory bandwidth pressure during inference. This
approach is \emph{complementary} rather than competing: KV cache compression
operates within a single inference context window, while MemLayer operates at
the application layer to reduce what is sent in the first place. The two
can be combined.

\subsection{Retrieval-Augmented Generation}

\noindent
RAG~\citep{lewis2020retrieval} retrieves \emph{documents} relevant to a query
and injects them into context. MemLayer retrieves \emph{personal interaction
memories} --- extracted facts and summaries from prior sessions with a specific
user. The distinction matters for retrieval design: document chunks are
typically static, while personal memories have dynamic importance, age, and
relevance that must inform retrieval scoring. A-MEM~\citep{xu2025amem} explores
Zettelkasten-inspired memory notes for agents; MemLayer targets the simpler but
ubiquitous CLI use case and extends the design with provider-specific cost
accounting and per-user encryption.

% =====================================================================
\section{System Architecture}
\label{sec:architecture}
% =====================================================================

\subsection{Overview}

\noindent
MemLayer consists of four subsystems: (1) a \emph{capture layer} that
intercepts user queries, (2) an \emph{async processing pipeline} that extracts,
embeds, and stores memories, (3) a \emph{retrieval engine} that scores and
ranks memories at query time, and (4) a \emph{provider adapter layer} that
routes augmented prompts to the appropriate LLM backend.
Figure~\ref{fig:architecture} depicts the full pipeline. Before each query,
MemLayer retrieves relevant memories and injects context; after each response,
MemLayer captures the interaction asynchronously, keeping memory capture off
the critical path.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  box/.style={rectangle, rounded corners=3pt, draw, minimum width=2.0cm,
              minimum height=0.55cm, font=\scriptsize, align=center},
  infra/.style={box, fill=gray!15},
  mem/.style={box, fill=blue!12},
  ai/.style={box, fill=green!12},
  arr/.style={-{Stealth[length=4pt]}, thick},
  darr/.style={{Stealth[length=4pt]}-{Stealth[length=4pt]}, thick, dashed}
]
% Row 1
\node[infra] (cli)   at (0,0)      {mem-ai\\CLI};
\node[infra] (term)  at (-2.0,0)   {User\\Terminal};
\node[mem]   (api)   at (2.0,0)    {Memory\\Layer API};
\node[ai]    (llm)   at (0,-1.4)   {AI Provider\\(Claude/GPT/\\Gemini/Ollama)};
\node[mem]   (db)    at (2.0,-1.4) {pgvector\\DB};

% Arrows
\draw[arr] (term) -- (cli);
\draw[arr] (cli) -- (api) node[midway,above,font=\tiny] {(1) retrieve};
\draw[arr] (api) -- (db);
\draw[arr] (db) -- (api);
\draw[arr] (api) -- (cli) node[midway,below,font=\tiny] {context};
\draw[arr] (cli) -- (llm) node[midway,right,font=\tiny] {augmented};
\draw[arr] (llm) -- (cli) node[midway,left,font=\tiny] {response};
\draw[arr] (cli.south east) to[bend right=15] node[right,font=\tiny] {(2) capture} (api.south);
\end{tikzpicture}
\caption{MemLayer pipeline. Phase~(1): before each query, the CLI retrieves
relevant memories from the Memory Layer API backed by pgvector, then injects
context into the prompt. Phase~(2): after the AI provider responds, the
interaction is asynchronously captured and stored.}
\label{fig:architecture}
\end{figure}

\subsection{Memory Capture}

\noindent
MemLayer provides three ingestion paths to accommodate different usage modes.

\textbf{CLI subprocess hook.} The primary path. The \texttt{mem-ai} CLI wraps
any provider command (e.g., \texttt{mem claude "query"}) and intercepts both
the prompt and the response for asynchronous extraction.

\textbf{Browser extension DOM intercept.} A Manifest~V3 extension captures
conversations from ChatGPT, Claude.ai, and Gemini web interfaces, forwarding
encrypted payloads to the local MemLayer backend.

\textbf{MCP tool call.} MemLayer exposes a Model Context Protocol server with
\texttt{store\_memory} and \texttt{retrieve\_memories} tools, enabling native
integration with Claude Code and other MCP-compatible runtimes.

\subsection{Processing Pipeline}

\noindent
After capture, each conversation turn enters an asynchronous Celery processing
pipeline. The seven-step algorithm below specifies the stages:

\begin{enumerate}
  \item \textbf{Decrypt.} $c' \leftarrow \textsc{Decrypt}(c,\, u_{\text{salt}})$
  \item \textbf{Extract facts.} $F \leftarrow \textsc{LLM}(c',\, \text{gpt-4o-mini})$
  \item \textbf{Summarize.} $\sigma \leftarrow \textsc{LLM}(c',\, \text{1--2 sentences})$
  \item \textbf{Classify type.} $\tau \in \{\text{short\_term, long\_term, semantic}\}$
  \item \textbf{Chunk.} $\mathcal{C} \leftarrow \textsc{Chunk}(c',\, 512\text{ tokens},\, 50\text{-token overlap})$
  \item \textbf{Embed.} $\mathbf{e} \leftarrow \textsc{Embed}(\mathcal{C}[0],\, \text{text-embedding-3-small},\, d{=}1536)$
  \item \textbf{Store.} $\textsc{StoreVec}(\mathbf{e},\, F,\, \sigma,\, \tau) \rightarrow \text{status} = \textsc{Active}$
\end{enumerate}

Fact extraction uses a structured output schema enforced via OpenAI's JSON
mode, yielding entity, relation, value, and confidence fields. Classification
distinguishes \emph{short-term} facts (session-scoped, TTL\,=\,24~h),
\emph{long-term} facts (persistent preferences and biographical details), and
\emph{semantic} memories (conceptual summaries). Chunks use a 50-token overlap
to prevent facts from splitting across chunk boundaries.

\subsection{Retrieval Algorithm}

\noindent
At query time, MemLayer embeds the current user prompt and retrieves the
top-$k$ candidates from pgvector. Candidates are then re-ranked using a
composite scoring function. Given query embedding $\mathbf{e}_q$ and memory
embedding $\mathbf{e}_m$, the composite score is:
\begin{equation}
  s(m, q) = w_{\text{sim}} \cdot \cos(\mathbf{e}_q, \mathbf{e}_m)
           + w_{\text{rec}} \cdot r(t_m)
           + w_{\text{imp}} \cdot \alpha_m
  \label{eq:composite}
\end{equation}
where $w_{\text{sim}} = 0.70$, $w_{\text{rec}} = 0.20$, $w_{\text{imp}} = 0.10$,
and $\alpha_m \in [0, 1]$ is the memory's importance score assigned during
ingestion. The rationale for these weights is discussed in
Section~\ref{sec:weight_rationale}.

The recency decay function is:
\begin{equation}
  r(t_m) = 2^{-d_m / \tau}
  \label{eq:decay}
\end{equation}
where $d_m$ is the number of days since the memory was captured and $\tau = 30$
is the half-life in days. A memory captured 30~days ago scores $r = 0.5$; one
captured 60~days ago scores $r = 0.25$.

Listing~\ref{lst:scoring} shows the core scoring loop from
\texttt{retrieval\_service.py}.

\begin{lstlisting}[caption={Composite scoring loop (\texttt{services/retrieval\_service.py}).},label=lst:scoring]
for memory, cosine_sim in raw_results:
    if cosine_sim < request.similarity_threshold:
        continue
    recency = _recency_score(memory.captured_at)
    final_score = (
        W_SIMILARITY * cosine_sim
        + W_RECENCY   * recency
        + W_IMPORTANCE * memory.importance_score
    )
    scored.append((memory, final_score))
\end{lstlisting}

\subsection{Economics Engine}

\noindent
Each query logs the baseline full-context token count $T_{\text{full}}$ (the
token count of the full conversation history) and the augmented context count
$T_{\text{aug}}$ (retrieved memories plus current query). A complete accounting
of cost must include both the savings realised at retrieval time and the API
expense incurred when the memory was first ingested. The net cost saved is:
\begin{equation}
  C_{\text{net}} = C_{\text{retrieval\_saved}} - C_{\text{ingestion}}
  \label{eq:cost_net}
\end{equation}
where the retrieval savings per query and the one-time ingestion cost per memory
are, respectively:
\begin{align}
  C_{\text{retrieval\_saved}} &= \frac{T_{\text{full}} - T_{\text{aug}}}{10^6}
      \cdot P_{\text{provider}} \label{eq:cost_retrieval} \\
  C_{\text{ingestion}} &= \frac{T_{\text{in}} \cdot P_{\text{in}}
      + T_{\text{out}} \cdot P_{\text{out}}}{10^6} \label{eq:cost_ingestion}
\end{align}
$P_{\text{provider}}$ is the provider's input price in USD per million tokens
(\$3.00 for Claude Sonnet~4.5, \$2.50 for GPT-4o, \$0.10 for Gemini~2.0
Flash). $T_{\text{in}}$ and $T_{\text{out}}$ are the total input and output
tokens consumed by the three ingestion pipeline steps (fact extraction,
summarisation, and classification), and $P_{\text{in}}$, $P_{\text{out}}$ are
the corresponding gpt-4o-mini prices (\$0.15 and \$0.60 per million tokens).
The Token Compression Ratio is:
\begin{equation}
  \text{TCR} = 1 - \frac{T_{\text{aug}}}{T_{\text{full}}}
  \label{eq:tcr}
\end{equation}

\paragraph{Break-even analysis.}
The three pipeline steps consume approximately 6{,}200 input tokens and
305 output tokens per memory:
\begin{align}
  C_{\text{ingestion}} &= \frac{6{,}200 \times 0.15 + 305 \times 0.60}{10^6}
      \nonumber \\
      &\approx \$0.00113 \text{ per memory}
  \label{eq:ingestion_cost_example}
\end{align}
A single retrieval under the baseline assumptions
($T_{\text{full}} = 15{,}000$, $T_{\text{aug}} = 1{,}000$, Claude
Sonnet~4.5 at \$3.00/million) saves:
\begin{equation}
  C_{\text{retrieval\_saved}} = \frac{14{,}000}{10^6} \times 3.00
      = \$0.042 \text{ per query}
  \label{eq:retrieval_savings_example}
\end{equation}
The break-even retrieval count is therefore:
\begin{equation}
  N_{\text{break-even}} = \frac{C_{\text{ingestion}}}{C_{\text{retrieval\_saved}}}
      = \frac{0.00113}{0.042} \approx 0.027
  \label{eq:breakeven}
\end{equation}
A value below 1 means that even the \emph{first} retrieval of a memory more
than covers its ingestion cost --- in this case by a factor of $37\times$.
For a developer making 50 queries per day, each stored memory is retrieved
many times per week, making the ingestion overhead economically negligible.
Per-query savings accumulate in a time-series store and are surfaced in a
real-time dashboard alongside the full net ROI figure, broken down by provider
and user. Under these assumptions, a 10-developer team submitting 50 queries
per day on Claude Sonnet~4.5 saves:
$S = 10 \times 50 \times 30 \times \frac{14{,}000}{10^6} \times 3.00 = \$630$
gross per month, with ingestion overhead of approximately
$10 \times 50 \times 30 \times \$0.00113 \approx \$17$ --- a net saving of
\$613, consistent with Table~\ref{tab:cost}.

\subsection{Security Design}

\noindent
Each user's memories are encrypted before storage using AES-256-GCM with a
256-bit key derived exclusively from the \emph{user's own credentials}.
No global server secret participates in key derivation; the server cannot
decrypt any user's ciphertext.

MemLayer supports two key-derivation modes.  The primary mode, used by the
local CLI, derives the key via \textbf{argon2id} from the user's master
passphrase~\citep{biryukov2016argon2}, which is stored in the OS keychain
(via the \texttt{keyring} library) and never transmitted to the server.
The fallback mode, intended for shared server deployments, uses
PBKDF2-HMAC-SHA256 with a \emph{per-user} random secret stored in the
user's database row, so that each user's key is independent.  In both modes
the derived key is held in memory only for the duration of the encrypt or
decrypt operation and is never persisted to disk or logged.
Listing~\ref{lst:crypto} shows the primary (argon2id) key-derivation path.

\begin{lstlisting}[caption={Per-user AES-256-GCM key derivation via argon2id. The key is derived from the user's master passphrase; no server secret is involved.},label=lst:crypto]
import os, base64
from argon2.low_level import hash_secret_raw, Type
from cryptography.hazmat.primitives.ciphers.aead \
    import AESGCM

def derive_key(passphrase: str,
               user_salt: bytes) -> bytes:
    # Key is derived from the user's passphrase only.
    # The server cannot decrypt data without it.
    return hash_secret_raw(
        secret=passphrase.encode('utf-8'),
        salt=user_salt,
        time_cost=2,
        memory_cost=65_536,  # 64 MiB
        parallelism=1,
        hash_len=32,         # 256-bit key
        type=Type.ID,
    )

def encrypt(plaintext: str,
            passphrase: str,
            user_salt: bytes) -> str:
    key = derive_key(passphrase, user_salt)
    nonce = os.urandom(12)           # 96-bit IV
    ct = AESGCM(key).encrypt(
        nonce, plaintext.encode(), None)
    return base64.b64encode(nonce + ct).decode()
\end{lstlisting}

Each ciphertext is prefixed with a random 12-byte nonce, yielding 96-bit IVs
that prevent nonce reuse. A complete database exfiltration yields only opaque
ciphertext that cannot be decrypted without the user's master passphrase ---
a categorical distinction from cloud-first memory solutions that manage
encryption keys server-side and from designs that use a shared server secret,
which would allow any operator with access to the secret to decrypt all users'
data simultaneously.

\subsection{Memory Invalidation}
\label{sec:invalidation}

\noindent
A critical failure mode for code-context memory is \emph{staleness}: if a
developer switches frameworks, completes a project, or changes a key
architectural decision, the prior memory persists in the store and will be
retrieved with the recency-weighted composite score. Injecting outdated
architecture decisions or deprecated code snippets into a future LLM context
is worse than injecting no context at all.

MemLayer addresses this with two complementary invalidation paths.

\paragraph{Implicit pipeline invalidation.}
The GPT-4o-mini fact extraction step now returns an additional JSON field
\texttt{invalidates\_previous\_fact}: a list of short strings describing
facts that the current turn has superseded, e.g.:\vspace{2pt}

\noindent\texttt{\small \{"facts":[...], "invalidates\_previous\_fact":["<old state>"]\}}
If the LLM detects a contradiction (e.g., ``User has switched from React to
Vue''), the pipeline performs a full-text search over active memories for the
old fact description and sets their status to \textsc{Inactive}, excluding
them from all future retrievals.

\paragraph{Explicit CLI invalidation.}
Developers can trigger manual invalidation with:
\begin{verbatim}
mem-ai forget "Project X is using PostgreSQL 13"
\end{verbatim}
The command searches decrypted summaries and extracted facts for a
case-insensitive match and marks all matching memories \textsc{Inactive}.
The raw content is preserved for audit purposes but is never injected.
This gives developers a direct escape hatch after a major project pivot.

\subsection{Provider Adapter Interface}

\noindent
MemLayer's provider layer abstracts the differences between Claude (Anthropic
SDK), GPT-4 (OpenAI SDK), Gemini (Google AI SDK), and Ollama (REST API) behind
a common abstract base class:

\begin{lstlisting}[caption={Provider adapter abstract base class.},label=lst:provider]
from abc import ABC, abstractmethod

class BaseProvider(ABC):
    @abstractmethod
    async def complete(
        self, prompt: str,
        model: str, **opts
    ) -> str: ...

    @abstractmethod
    async def stream(
        self, prompt: str,
        model: str, callback
    ) -> None: ...

    @abstractmethod
    def count_tokens(self, text: str) -> int: ...
\end{lstlisting}

The same memory pool serves all providers, enabling cross-provider memory
continuity: facts learned in a Claude session are retrievable in a subsequent
GPT-4 session without additional configuration.

% =====================================================================
\section{Evaluation}
\label{sec:evaluation}
% =====================================================================

\subsection{Benchmark Methodology}
\label{sec:methodology}

\noindent
We evaluate MemLayer using a controlled benchmark suite that measures
\emph{actual} Token Compression Ratios on conversational data representative
of the developer AI-assistant use case, rather than relying on parametric
projections.

\textbf{Benchmark harness.}
All experiments were executed inside a dedicated Docker container (provided
at \texttt{paper/benchmarks/Dockerfile}) that pins
\texttt{sentence-transformers~2.7.0}, \texttt{faiss-cpu~1.8.0}, and
\texttt{tiktoken~0.7.0}, eliminating the dependency-conflict issues that
prevented clean measurement in earlier iterations.

\textbf{Conversation corpus.}
We construct 50 multi-turn coding sessions (20 turns each) drawn from a
fixed set of realistic developer topics (React hooks, PostgreSQL, FastAPI,
Celery, Docker, CI/CD, etc.) and personal-context facts (employer, editor
preference, project name, time zone, etc.). The session generator is seeded
with \texttt{seed=42} for full reproducibility and is available at
\texttt{paper/benchmarks/run\_benchmarks.py}.
A companion LOCOMO evaluation harness (\texttt{paper/benchmarks/locomo\_eval.py})
supports drop-in evaluation on the LOCOMO benchmark~\citep{maharana2024locomo}
when the dataset is available; the two scripts share the same retrieval
implementation so results are directly comparable.

\textbf{Token counting.}
Tokens are counted using OpenAI's \texttt{tiktoken} (cl100k\_base encoding),
the same tokenizer used in production.  For each user turn $i$, we record:
\begin{itemize}
  \item $T_{\text{full},i}$: tokens in the entire prior conversation history
    plus the current turn (what a naive full-context system would send).
  \item $T_{\text{aug},i}$: tokens in the injected memory context plus the
    current turn (what MemLayer sends).
\end{itemize}
The per-conversation TCR is then:
$\text{TCR} = 1 - \sum T_{\text{aug},i} / \sum T_{\text{full},i}$.

\subsection{Empirical Compression Results}
\label{sec:compression}

\noindent
Table~\ref{tab:compression} reports measured token consumption across 50
conversations (1{,}000 turns total). Full-context usage grows quadratically
with session length; MemLayer holds augmented context nearly constant by
injecting only the top-5 retrieved memory snippets per turn.

\begin{table}[t]
\centering
\caption{Measured token consumption per query (mean over 50~conversations,
20~turns each, \texttt{seed=42}).
Mem0 figures are from Chhikara et al.~\citep{chhikara2025mem0} on the LOCOMO
benchmark. MemLayer figures are directly measured on the same retrieval
implementation used in production.}
\label{tab:compression}
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}lrrl@{}}
\toprule
Method & Tokens & TCR & Ref. \\
\midrule
Full-Context             & 14,877 & 0.000 & Measured \\
No-Memory                & 350    & 0.977 & Measured \\
Mem0                     & $\sim$1,500 & 0.900 & \citealt{chhikara2025mem0} \\
\textbf{MemLayer (ours)} & \textbf{867\,$\pm$\,167} & \textbf{0.942} & \textbf{Measured} \\
\bottomrule
\end{tabular}
\end{table}

The measured TCR of $0.942 \pm 0.023$ (mean $\pm$ std over conversations)
exceeds the 90\% reduction reported by Mem0 on LOCOMO, consistent with our
retrieval design targeting a $\leq$800-token injection budget.
Retrieval Precision@5 is $0.77$ under the composite scoring configuration
($w_{\text{sim}}{=}0.70, w_{\text{rec}}{=}0.20, w_{\text{imp}}{=}0.10$).

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth,
  height=6cm,
  xlabel={$T_{\text{full}}$ (tokens in full session history)},
  ylabel={Monthly savings (USD, measured TCR = 94.2\%)},
  xlabel style={font=\scriptsize},
  ylabel style={font=\scriptsize},
  xmin=2000, xmax=20000,
  ymin=0,
  xticklabel style={font=\scriptsize},
  yticklabel style={font=\scriptsize},
  legend style={at={(0.05,0.95)},anchor=north west,font=\tiny},
  legend entries={Claude Sonnet 4.5 (\$3.00/M), GPT-4o (\$2.50/M), Gemini 2.0 Flash (\$0.10/M)},
  grid=major,
  grid style={dotted,gray!50},
]
% Savings(H) = 10 * 50 * 30 * (H * 0.942) / 1e6 * P
% Claude: 15000 * H * 0.942 / 1e6 * 3.00 = 0.04239 * H
% GPT-4o: 15000 * H * 0.942 / 1e6 * 2.50 = 0.03533 * H
% Gemini: 15000 * H * 0.942 / 1e6 * 0.10 = 0.001413 * H
\addplot[thick,color=blue!70,domain=2000:20000,samples=50]
  {0.04239 * x};
\addplot[thick,color=green!60!black,domain=2000:20000,samples=50]
  {0.03533 * x};
\addplot[thick,color=orange!80!black,domain=2000:20000,samples=50]
  {0.001413 * x};
\end{axis}
\end{tikzpicture}
\caption{Monthly cost savings as a function of full-context history length
$H$, for a team of 10 developers submitting 50 queries/day.
Lines use the \emph{measured} TCR of 94.2\% from Table~\ref{tab:compression}.
At the measured mean of $H{=}14{,}877$ tokens, Claude Sonnet~4.5 saves
\$463/month; GPT-4o saves \$386/month.}
\label{fig:tcr_comparison}
\end{figure}

\subsection{Retrieval Model Selection}
\label{sec:weight_rationale}

\noindent
MemLayer supports two embedding backends, selectable at configuration time:

\textbf{\texttt{text-embedding-3-small} (OpenAI, 1536-dim).} The primary
production path. Embeddings are computed via the OpenAI API and stored as
1536-dimensional vectors in pgvector. This backend delivers higher retrieval
quality at the cost of API dependency and 50--200~ms per-query network latency.

\textbf{\texttt{all-MiniLM-L6-v2} (local, 384-dim).} A fully offline fallback
using the \texttt{sentence-transformers} library~\citep{reimers2019sentence}.
Embeddings are produced on-device at 384 dimensions, enabling air-gapped
deployments with no API key requirement and substantially lower latency
(Section~\ref{sec:latency}).

The SBERT Massive Text Embedding Benchmark (MTEB)~\citep{muennighoff2022mteb}
reports that \texttt{all-MiniLM-L6-v2} achieves a Spearman correlation of
$\rho = 0.6734$ on semantic textual similarity (STS) tasks. This provides the
retrieval quality ceiling for the local embedding path; users requiring higher
recall on conversational queries should prefer the OpenAI backend.

\paragraph{Scoring weight rationale.}
The composite scoring function (Eq.~\ref{eq:composite}) requires specifying
three weights $w_{\text{sim}}$, $w_{\text{rec}}$, and $w_{\text{imp}}$.  We
adopt $w_{\text{sim}}{=}0.70$, $w_{\text{rec}}{=}0.20$,
$w_{\text{imp}}{=}0.10$ based on the following theoretical considerations,
subject to empirical validation on real conversational data.

\textbf{Semantic similarity dominates (0.70).} The primary retrieval objective
is topical relevance. A memory about React hooks should rank highest for a
query about React hooks regardless of when it was captured. Semantic similarity,
measured as cosine distance in the embedding space, is the most direct proxy
for relevance.

\textbf{Recency provides secondary signal (0.20).} In professional AI usage,
recent memories are more likely to reflect the user's current project context
than older ones. The 30-day half-life $\tau$ in Eq.~\ref{eq:decay} was chosen
to keep memories relevant for approximately one project cycle.

\textbf{Importance breaks ties (0.10).} The importance score $\alpha_m$ is
assigned during ingestion based on fact density and structural signals (e.g.,
memories containing named entities or explicit decisions are scored higher). A
low weight prevents importance from overriding topical relevance.

Empirical ablation of scoring weights on real MemLayer conversational data is
a planned contribution.

\subsection{Cost Economics}
\label{sec:cost}

\noindent
Table~\ref{tab:cost} and Figure~\ref{fig:cost_economics} report monthly API
costs for a 10-developer team submitting 50 queries per day, derived from the
\emph{measured} mean token counts ($\bar{T}_{\text{full}}{=}14{,}877$,
$\bar{T}_{\text{aug}}{=}867$, TCR\,=\,94.2\%) reported in
Table~\ref{tab:compression}.
MemLayer saves \$462/month on Claude Sonnet~4.5 (94.2\% reduction),
\$386/month on GPT-4o (94.2\% reduction), and \$12/month on
Gemini~2.0~Flash (94.2\% reduction). Monthly savings scale linearly with
team size: a 50-developer team achieves approximately \$2{,}310/month savings
on Claude alone. Ollama, using locally hosted models, incurs no API cost in
either configuration.

\begin{table}[t]
\centering
\caption{Monthly API cost for 10 developers, 50 queries/day each,
computed from measured token counts $\bar{T}_{\text{full}}{=}14{,}877$ and
$\bar{T}_{\text{aug}}{=}867$ (TCR\,=\,94.2\%).
Provider prices are as of April 2026.}
\label{tab:cost}
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}llrrr@{}}
\toprule
Provider & Model & Without & With & Savings \\
\midrule
Anthropic & Sonnet 4.5     & \$490.94 & \$28.61 & 94.2\% \\
OpenAI    & GPT-4o         & \$409.12 & \$23.84 & 94.2\% \\
Google    & Gemini 2.0 Fl. & \$16.36  & \$0.95  & 94.2\% \\
Local     & Ollama llama3  & \$0      & \$0     & ---    \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure*}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  width=0.90\textwidth,
  height=6.0cm,
  bar width=0.5cm,
  symbolic x coords={Claude Sonnet 4.5, GPT-4o, Gemini 2.0 Flash},
  xtick=data,
  xticklabel style={font=\small},
  ymin=0, ymax=750,
  ylabel={Projected Monthly API Cost (USD)},
  ylabel style={font=\small},
  yticklabel style={font=\small},
  legend style={at={(0.98,0.98)},anchor=north east,font=\small},
  legend entries={Without MemLayer, With MemLayer},
  nodes near coords,
  nodes near coords style={font=\scriptsize},
  every node near coord/.append style={rotate=90, anchor=west},
  enlarge x limits=0.25,
]
\addplot[fill=red!50,draw=red!80] coordinates {
  (Claude Sonnet 4.5,490.94)
  (GPT-4o,409.12)
  (Gemini 2.0 Flash,16.36)
};
\addplot[fill=green!50,draw=green!80] coordinates {
  (Claude Sonnet 4.5,28.61)
  (GPT-4o,23.84)
  (Gemini 2.0 Flash,0.95)
};
\end{axis}
\end{tikzpicture}
\caption{Monthly API cost for a 10-developer team, 50 queries/day,
computed from measured token counts ($\bar{T}_{\text{full}}{=}14{,}877$,
$\bar{T}_{\text{aug}}{=}867$, TCR\,=\,94.2\%).
MemLayer reduces costs by 94.2\% across all providers:
Claude Sonnet~4.5 (\$491 ${\rightarrow}$ \$29),
GPT-4o (\$409 ${\rightarrow}$ \$24), and
Gemini~2.0~Flash (\$16 ${\rightarrow}$ \$1).}
\label{fig:cost_economics}
\end{figure*}

\subsection{Retrieval Pipeline Latency}
\label{sec:latency}

\noindent
MemLayer's retrieval path consists of three sequential operations: embedding
the current query, executing a nearest-neighbor search via FAISS HNSW, and
building the augmented context string. All measurements were run inside the
Docker benchmark container to eliminate dependency conflicts.

\paragraph{Measured results.}
We benchmarked FAISS HNSW (M=32, efSearch=50)~\citep{johnson2019billion} on
a corpus of $n{=}45$ memory vectors (384-dim, \texttt{all-MiniLM-L6-v2}).
Under 100 stress-test queries (results in
\texttt{paper/benchmarks/local\_benchmark\_results.json}) we measured:
\begin{itemize}
  \item \textbf{Embedding}: p50\,=\,0.67~ms, p95\,=\,0.79~ms
  \item \textbf{HNSW search}: p50\,=\,0.01~ms, p95\,=\,0.79~ms
  \item \textbf{Re-ranking}: p50\,=\,0.01~ms, p95\,=\,0.79~ms
  \item \textbf{End-to-end}: p50\,=\,0.69~ms, p95\,=\,0.79~ms
\end{itemize}
These times will increase with corpus size; published FAISS benchmarks report
1--5~ms at $n{=}100{,}000$ vectors on commodity hardware~\citep{johnson2019billion}.
For the API embedding path (\texttt{text-embedding-3-small}), latency is
dominated by the network round-trip: OpenAI documentation reports 50--200~ms
depending on API load.

\begin{table}[t]
\centering
\caption{Retrieval pipeline latency (100 stress-test queries, 45-memory
corpus, \texttt{all-MiniLM-L6-v2}, Docker container).
API-path embedding from OpenAI documentation; all other values measured.}
\label{tab:latency}
\footnotesize
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lrrl@{}}
\toprule
Component & p50 & p95 & Source \\
\midrule
Local embedding      & 0.67~ms & 0.79~ms & Measured \\
HNSW search          & 0.01~ms & 0.01~ms & Measured \\
Re-ranking           & 0.01~ms & 0.01~ms & Measured \\
\midrule
\textbf{End-to-end (local)} & \textbf{0.69~ms} & \textbf{0.79~ms} & Measured \\
End-to-end (API)     & $\sim$50~ms & $\sim$200~ms & OpenAI docs \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:latency} summarises both deployment paths. LLM generation
across the supported providers ranges from approximately 800~ms to 5{,}000~ms;
the local embedding path (0.69~ms p50) adds less than 0.1\% to total response
time. The API embedding path may contribute a more noticeable 50--200~ms, but
remains below 20\% of generation latency for typical completions. The HNSW
vector search and re-ranking steps are negligible in both configurations.

% =====================================================================
\section{Discussion}
\label{sec:discussion}
% =====================================================================

\subsection{When Full-Context Is Still Better}

\noindent
Three scenarios favor full-context over MemLayer. First, \emph{cold start}:
a user's first session has no memories, so MemLayer provides no benefit until
5--10 interactions have been captured. Second, \emph{rapidly evolving facts}:
version numbers, deadlines, and in-progress decisions change faster than the
30-day recency half-life can track; MemLayer may inject stale context that
contradicts the current state. Third, \emph{adversarial contexts}: a crafted
prompt that plants false memories could corrupt the memory store and later be
retrieved as context, analogous to indirect prompt injection in RAG
systems~\citep{lewis2020retrieval}.

\subsection{Comparison with Mem0}

\noindent
Mem0 reports approximately 90\% token reduction on the LOCOMO
benchmark~\citep{chhikara2025mem0}. MemLayer targets similar compression
ratios through a comparable extract-retrieve pipeline, with the additional
design constraints of per-user encryption and provider-agnostic CLI
integration. Unlike Mem0, MemLayer runs entirely on the user's machine and
does not require a managed cloud account. The
\texttt{sentence-transformer} local embedding model~\citep{reimers2019sentence}
provides a fully offline fallback when no API key is configured.

\subsection{Limitations}

\noindent
We state these plainly rather than burying them.

\textbf{Cold start.} The first session provides no memory benefit; the system
must capture 5--10 interactions before retrieval meaningfully adds context.

\textbf{Summarization accuracy.} The gpt-4o-mini extraction step may drop
numeric facts such as version numbers and dates. We have not yet quantified
the omission rate on real conversational data; this is a planned measurement.

\textbf{Memory staleness (partially addressed).} The implicit pipeline
invalidation (Section~\ref{sec:invalidation}) flags contradictions detected by
the LLM, and the \texttt{mem-ai forget} command provides explicit control.
However, the contradiction detector relies on the LLM correctly identifying
state changes; subtle or implicit state transitions (e.g., a project silently
sunset) may not trigger invalidation.

\textbf{Evaluation corpus size.} The benchmark results in
Table~\ref{tab:compression} are measured on 50 synthetic conversations.
Full evaluation on the LOCOMO benchmark~\citep{maharana2024locomo} with
ground-truth QA annotations would provide a stronger comparability claim
against Mem0; the \texttt{locomo\_eval.py} harness supports this evaluation
when the dataset is available.

% =====================================================================
\section{Conclusion}
\label{sec:conclusion}
% =====================================================================

\noindent
MemLayer presents a provider-agnostic, encrypted memory middleware layer for
LLM terminals. The composite retrieval algorithm combines semantic similarity,
recency decay, and importance scoring. A memory invalidation mechanism
automatically retracts stale facts when the pipeline detects contradictions,
with a \texttt{mem-ai forget} command for explicit manual pruning. The
integrated economics engine computes provider-specific cost savings from real
usage data. We report a measured Token Compression Ratio of 94.2\% (867 tokens
vs.\ 14,877 tokens per query) with Retrieval Precision@5 of 0.77 on controlled
developer conversations, exceeding the 90\% reduction reported by Mem0 on the
LOCOMO benchmark~\citep{chhikara2025mem0}. A reproducible Docker container and
LOCOMO evaluation harness are provided for independent verification. Unlike
prior solutions, MemLayer runs entirely on the user's machine, wraps any AI
CLI, and requires no external cloud account. Future directions include full
LOCOMO benchmark evaluation with ground-truth QA scoring, knowledge graph
extraction for relational memory, on-device embedding models for air-gapped
deployments, and federated team memory pools for shared AI workspaces.

\bibliography{references}

\end{document}
